The brightness and spatial distributions of terrestrial radio sources

### 2013-01-07 articles/Article-RFI-distribution/article-rfi-distribution.tex
Stefan's comments part I

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are coherently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, we conclude that it is probable that most of the RFI sources do not add coherently.
\end{abstract}

@@

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise, to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) \citep{de-bruyn-eor-ursi-2011}.

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions. This results in a better insight in the effects of low-level interference and allows one to simulate the effects of interference more accurately. A distribution can be expressed either as a differential or cumulative function. This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.

@@

\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth). For example, aeroplanes and satellites have widely different heights, while other sources are ground-based. Even ground-based sources might be spread differently. For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are distributed differently from broadcasting transmitters. The frequencies of deliberately transmitting sources of a particular class of devices are constrained by the bands that have been allocated for the given class, which might allow one to distinguish different sources to some extent.

@@

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane. Therefore, RFI follows a power-law distribution.

@@

\subsection{Creating a histogram} \label{sec:histogram}
While creating a histogram is trivial, it is important to note that it is necessary to have a variable bin size. This is mandated by the large dynamic range of the histogram that we are interested in. Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the frequency distribution is normalized by the bin size after counting. Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot. This can be seen in the tails of the distributions of Fig.~\ref{fig:rayleigh-and-rfi-combined}. This, however, is not an intrinsic feature of the data but a consequence of the binning method.

@@

One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{max}} \right)^{-1}.
\end{equation}
However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution. Moreover, the estimator requires an iteration over all data points, and because of the size of the data sets this is somewhat impractical. However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ the bin frequency. Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator. This avoids the need to iterate over all data.

@@

It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying source populations. Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands. Nevertheless, combining the data of all the sub-bands results in reasonably-stable power-law distributions. This indicates that the individual sub-bands have an approximately similar underlying power-law distribution, that is not yet apparent because not enough samples are combined in the histograms of individual sub-bands.

@@

To make sure that the antenna response does not influence the result of the slope, we have also analysed the curves after a simple band-pass calibration. This was performed by dividing each sub-band by its standard deviation after RFI excision. Because the standard deviation of the distribution might be affected by the RFI tail of the distribution, we compare the two histograms to make sure the power-law distribution is not significantly changed. The resulting histograms are shown in Fig.~\ref{fig:histogram-passband-corrected}. This procedure makes the bulge of the LBA histogram similar to the bulge of a Rayleigh curve and extends the power-law part. Nevertheless, it does not change the log-log slope of the power law in either histograms. This validates that the variable gain that is caused by the antenna response does not change the observed power law. Consequently, it can be expected that other stochastic effects, such as the intrinsic source strength and the beam gain due to a differing direction of arrival, will similarly not effect the power law. Because the band-pass corrected histograms provide a more representative analysis, we will use the corrected histograms for further analysis.

@@

The Rayleigh parts of the distributions are plotted in Fig.~\ref{fig:histograms-with-fits}, along with a least-squares fit and its residuals. Both histograms follow the Rayleigh distribution for about five orders of magnitude, which is validated by the residuals that show only noise. It breaks down about one order of magnitude before the mode of the distributions. This is because of the multi-variate nature of the distributions, as was described in \S\ref{sec:histogram-noise}.

@@

\subsection{Interference-reducing effects} \label{sec:coherence-reduction}
When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution, before further averaging. There are several reasons for this, which we will summarize one by one.
\begin{itemize}
 \item Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes. This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise.
 \item Many RFI sources will be seen by only a few stations. This would make the histogram go down more quickly at lower amplitudes.
 \item For the shortest baselines at 150~MHz, the far field starts around 1~km. Some RFI sources will therefore be in the near field, especially for the longer baselines. In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see them with different phases.
 \item Many stationary sources are not constant over time, thus will be attenuated somewhat by averaging.
 \item We have assumed 100\% of the spectrum is occupied by RFI. If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases with about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much the electromagnetic spectrum is truly occupied.
 \item Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other. Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread. Therefore, they will add incoherently.
\end{itemize}

@@

With the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still remains to be seen. To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels. The current results indicate that a lot of RFI is not coherent, and the situation is promising. Considering the current RFI results, and the availability of further mitigation steps, we conclude that RFI will likely not be problematic for the detection Epoch of Reionisation with LOFAR.

>>>>

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and coherently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, we conclude that it is probable that most of the RFI sources do not add coherently.
\end{abstract}

@@

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise, to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) \citep{de-bruyn-eor-ursi-2011}.

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions. This results in a better insight into the effects of low-level interference and allows one to simulate the effects of interference more accurately. A distribution can be expressed either as a differential or cumulative function. This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.

@@

\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth). For example, aeroplanes and satellites have widely different heights, while other sources are ground-based. Even ground-based sources might be spread differently. For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are distributed differently from broadcasting transmitters. For deliberate transmitters, the frequency at which interference occurs can identify the involved class of devices, because devices are constrained by the bands that have been allocated for the given class.

@@

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.

@@

\subsection{Creating a histogram} \label{sec:histogram}
While creating a histogram is trivial, it is important to note that it is necessary to have a variable bin size. This is mandated by the large dynamic range of the histogram that we are interested in. Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the frequency counts are divided by the bin size after counting. Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot. This can be seen in the tails of the distributions of Fig.~\ref{fig:rayleigh-and-rfi-combined}. This, however, is not an intrinsic feature of the data but a consequence of the binning method.

@@

One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{max}} \right)^{-1}.
\end{equation}
However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution. Moreover, the estimator requires reading and processing all data points, and because of the size of the data sets this is somewhat impractical. However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ the bin frequency. Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator. This avoids the need to iterate over all data.

@@

It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying source populations. Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands. Nevertheless, combining the data of all the sub-bands results in reasonably stable power-law distributions. This indicates that the individual sub-bands have an approximately similar underlying power-law distribution, that is not yet apparent because not enough samples are combined in the histograms of individual sub-bands.

@@

To make sure that the antenna response does not influence the result of the slope, we have also analysed the curves after a simple band-pass calibration. This was performed by dividing each sub-band by its standard deviation after RFI excision. Because the standard deviation of the distribution might be affected by the RFI tail of the distribution, we compare the two histograms to make sure the power-law distribution is not significantly changed. The resulting histograms are shown in Fig.~\ref{fig:histogram-passband-corrected}. This procedure makes the bulge of the LBA histogram similar to the bulge of a Rayleigh curve and extends the power-law part. Nevertheless, it does not change the log-log slope of the power law in either histograms. This validates that the variable gain that is caused by the antenna response does not change the observed power law. Consequently, it can be expected that other stochastic effects, such as the intrinsic source strength and the beam gain due to a differing direction of arrival, will similarly not affect the power law. Because the band-pass corrected histograms should provide a more accurate analysis, we will use the corrected histograms for further analysis.

@@

The Rayleigh parts of the distributions are plotted in Fig.~\ref{fig:histograms-with-fits}, along with a least-squares fit and its residuals. Both histograms follow the Rayleigh distribution for about five orders of magnitude, which is validated by the residuals that show only noise. It breaks down about one order of magnitude before the mode of the distributions. This is because of the multi-component nature of the distributions, as was described in \S\ref{sec:histogram-noise}.

@@

\subsection{Interference-reducing effects} \label{sec:coherence-reduction}
When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution, before further averaging. There are several reasons for this, which we will summarize one by one.
\begin{itemize}
 \item Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes. This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise.
 \item Many RFI sources will be seen by only a few stations. This would make the histogram go down more quickly at lower amplitudes.
 \item For the shortest baselines at 150~MHz, the far field starts around 1~km. Some RFI sources will therefore be in the near field, especially for the longer baselines. In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see them with different phases.
 \item Many stationary sources are not constant over time, thus will be attenuated somewhat by averaging.
 \item We have assumed 100\% of the spectrum is occupied by RFI. If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.
 \item Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other. Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread. Therefore, they will add incoherently.
\end{itemize}

@@

With the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still remains to be seen. To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels. The current results indicate that a lot of RFI is not coherent, and the situation is promising. Considering the current RFI results, and the availability of further mitigation steps, we conclude that RFI will likely not be problematic for the detection of the Epoch of Reionisation with LOFAR.
